A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-20 tex/body.tex
some fixes

Formally, $\mu$-\emph{terms} are defined recursively as being either a variable $x$
or an applied constructor
$\MU x.\: \const{C}( \vec t\, )$ for some $\const{C} \in \Ctr$ and
$\mu$-terms $\vec t$ of the expected types.
The variable $x$ need not occur freely in the $\mu$-binder's body, in which case
the binder can be omitted.
In the \rn{Acyclic} rule,
$\FV( u )$ denotes the variables occurring freely in~$u$.
%%% But there are no constants of type $\tau \in \Nondata$ -- at most variables!
%
%For uniformity, we also consider $%\MU x.\;
%\const{c}$ to be a $\mu$-term if $\const{c}$ is a constant of type $\tau \in \Nondata$.
%
%%% I think this should be clear enough from the "Conventions" paragraph at the
%%% end of the introduction.
%%
% which we define recursively:
%if $\const{C} : \vec \tau \rightarrow \upsilon$
%and $\vec v$ are well-typed $\mu$-terms of type $\vec \tau$ under the assumption that $x$ has type $\upsilon$,
%then $\MU x.\; \const{C}( \vec v )$ is a well-typed term of type $\upsilon$.
%
A $\mu$-term is \emph{closed} if %and only if --- by convention, "and only if" is not necessary for *definitions* (but add it back if you disagree)
it contains no free variables. \,It is \emph{cyclic} if %and only if --- ditto
it contains a bound variable. \,The
\goodbreak %%% TYPESETTING: horrible hacks (including two \,'s)
\noindent
$\alpha$-\emph{equivalence} relation $t \aequiv u$
indicates that the $\mu$-terms $t$ and $u$
are syntactically equivalent for some capture-avoiding renaming of $\mu$-bound variables\begin{rep}---e.g.,
%$\MU x.\; \const{C}( x ) \aequiv \MU y.\; \const{C}( y )$ and
$\MU x.\; \const{D}( y, x ) \aequiv \MU z.\; \const{D}( y, z )$,
but
$\MU x.\; \const{C}( x ) \not\aequiv \MU x.\; \const{D}( y, x ) \not\aequiv \MU x.\; \const{D}( z, x )
\not\aequiv \MU y.\; \const{D}( y, x )$\end{rep}.
Two $\mu$-terms can denote the same value despite being $\alpha$-disequivalent---e.g.,
$\MU x.\; \const{C}( x ) \not\aequiv \MU y.\; \const{C}( \const{C}( y ) )$.
%, or informally are \emph{observationally equivalent}.
%For convenience, we use $\mu$-terms to refer to (classes of) values for both codatatype and datatype terms,
%where in the latter case, a $\mu$-term with a bound variable denotes an infeasible term.
%where the latter case adds the restriction on $\mu$-terms $t$ that no subterms of $t$ contain are bound variables.

@@

Before constructing a model for $F\!$, the theory solver constructs neither $\mu$-terms nor the mapping $\Val$.
Instead, $\Val$ is computed implicitly by traversing the equivalence classes of $\Ec$ during phase~2.
To detect whether \rn{Acyclic} applies,
the procedure considers each equivalence class $\ec{t}$ containing a datatype constructor $\const{C}( t_1, \ldots, t_n )$.
It visits $\ec{t_1}, \ldots, \ec{t_n}$ and all constructor arguments in these equivalence classes recursively.
If while doing so it returns to $\ec{t}$, it deduces that \rn{Acyclic} applies. % to $t$.
To recognize when the precondition of \rn{Unique} holds,
the procedure considers the set $\SSS$ of all codatatype equivalence classes.
It
%
\begin{conf}\goodbreak\noindent\end{conf}%
simultaneously visits the equivalence classes of arguments of constructor terms in each equivalence class in $\SSS$,
while partitioning $\SSS$ into $\SSS_1, \ldots, \SSS_n$ based on the top-most symbol of constructor terms in these equivalence classes
and the equivalence of their arguments of ordinary types.
It then partitions each set %in this partition %in the same manner
recursively.
If %after doing so,
the resulting partition contains a set $\SSS_i$ containing distinct terms $u$
and $v$, it deduces that \rn{Unique} applies to $u$ and~$v$.

>>>>

Formally, $\mu$-\emph{terms} are defined recursively as being either a variable $x$
or an applied constructor
$\MU x.\: \const{C}( \vec t\, )$ for some $\const{C} \in \Ctr$ and
$\mu$-terms $\vec t$ of the expected types.
The variable $x$ need not occur freely in the $\mu$-binder's body, in which case
the binder can be omitted.
In the \rn{Acyclic} rule,
$\FV( u )$ denotes the variables occurring freely in~$u$.
%%% But there are no constants of type $\tau \in \Nondata$ -- at most variables!
%
%For uniformity, we also consider $%\MU x.\;
%\const{c}$ to be a $\mu$-term if $\const{c}$ is a constant of type $\tau \in \Nondata$.
%
%%% I think this should be clear enough from the "Conventions" paragraph at the
%%% end of the introduction.
%%
% which we define recursively:
%if $\const{C} : \vec \tau \rightarrow \upsilon$
%and $\vec v$ are well-typed $\mu$-terms of type $\vec \tau$ under the assumption that $x$ has type $\upsilon$,
%then $\MU x.\; \const{C}( \vec v )$ is a well-typed term of type $\upsilon$.
%
A $\mu$-term is \emph{closed} if %and only if --- by convention, "and only if" is not necessary for *definitions* (but add it back if you disagree)
it contains no free variables. It is \emph{cyclic} if %and only if --- ditto
it contains a bound variable. The $\alpha$-\emph{equivalence} relation $t \aequiv u$
indicates that the $\mu$-terms $t$ and $u$
are syntactically equivalent for some capture-avoiding renaming of $\mu$-bound variables\begin{rep}---e.g.,
%$\MU x.\; \const{C}( x ) \aequiv \MU y.\; \const{C}( y )$ and
$\MU x.\; \const{D}( y, x ) \aequiv \MU z.\; \const{D}( y, z )$,
but
$\MU x.\; \const{C}( x ) \not\aequiv \MU x.\; \const{D}( y, x ) \not\aequiv \MU x.\; \const{D}( z, x )
\not\aequiv \MU y.\; \const{D}( y, x )$\end{rep}.
Two $\mu$-terms can denote the same value despite being $\alpha$-disequivalent---e.g.,
$\MU x.\; \const{C}( x ) \not\aequiv \MU y.\; \const{C}( \const{C}( y ) )$.
%, or informally are \emph{observationally equivalent}.
%For convenience, we use $\mu$-terms to refer to (classes of) values for both codatatype and datatype terms,
%where in the latter case, a $\mu$-term with a bound variable denotes an infeasible term.
%where the latter case adds the restriction on $\mu$-terms $t$ that no subterms of $t$ contain are bound variables.

@@

Before constructing a model for $F\!$, the theory solver constructs neither $\mu$-terms nor the mapping $\Val$.
Instead, $\Val$ is computed implicitly by traversing the equivalence classes of $\Ec$ during phase~2.
To detect whether \rn{Acyclic} applies,
the procedure considers each equivalence class $\ec{t}$ containing a datatype constructor $\const{C}( t_1, \ldots, t_n )$.
It visits $\ec{t_1}, \ldots, \ec{t_n}$ and all constructor arguments in these equivalence classes recursively.
If while doing so it returns to $\ec{t}$, it deduces that \rn{Acyclic} applies. % to $t$.
To recognize when the precondition of \rn{Unique} holds,
the procedure considers the set $\SSS$ of all codatatype equivalence classes. It
\begin{conf}\goodbreak\noindent\end{conf}%
simultaneously visits the equivalence classes of arguments of constructor terms in each equivalence class in $\SSS$,
while partitioning $\SSS$ into $\SSS_1, \ldots, \SSS_n$ based on the top-most symbol of constructor terms in these equivalence classes
and the equivalence of their arguments of ordinary types.
It then partitions each set %in this partition %in the same manner
recursively.
If %after doing so,
the resulting partition contains a set $\SSS_i$ containing distinct terms $u$
and $v$, it deduces that \rn{Unique} applies to $u$ and~$v$.
